---
site: bookdown::bookdown_site
output:
  html_document:
    df_print: paged
  bookdown::pdf_book:
    keep_tex: yes
  bookdown::gitbook:
    lib_dir: book_assets
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)

```

# Analysis of Time-to-Event Data: Study Sheet 2

# Submission: René-Marcel Kruse, Doctorial Candidate


# Exercise 1:
_Freireich et al. (1963) report the results of a clinical trial of a drug 6-mercaptopurine (6-
MP) versus a placebo in 42 children with acute leukemia. The trial was conducted at 11
American hospitals. Patients were selected who had a complete or partial remission of their
leukemia induced by treatment with the drug prednisone. (A complete or partial remission
means that either most or all signs of disease had disappeared from the bone marrow.)
The trial was conducted by matching pairs of patients at a given hospital by remission
status (complete or partial) and randomizing within the pair to either a 6-MP or placebo
maintenance therapy. Patients were followed until their leukemia returned (relapse) or until
the end of study. The data are given in the data frame drug6mp in the R package `KMsurv`. Make
use of `help(drug6mp)` to become acquainted with the data set. Note that the remission times
in Freireich et al. (1963) were measured in weeks and not, as indicated in `help(drug6mp)`, in
months!_

## (a)
_Consider only the 6-MP patients. Compute (using only your pocket calculator) the
Kaplan-Meier estimate of the survivor function along with the standard errors._ 

Kaplan-Meier Estimator given as:
$$\hat{S}(t) = \prod_{t_{k} \leq t} (1 - \dfrac{d_{k}}{n_{k}})$$
Variance of KP Estimator is as follows:
$$ var(\hat{S}(t)) = \hat{S}(t)^{2} \cdot \sum_{t_{k} \leq t} \dfrac{d_{k}}{n_{k}(n_{k} - d_{k})})$$

| $t_{k}$ | $n_{k}$ | $d_{k}$ | $\dfrac{d_{k}}{n_{k}}$      | $(1 - \dfrac{d_{k}}{n_{k}})$    | $\hat{S}(t)$                                                        | $\dfrac{d_{k}}{n_{k}(n_{k} - d_{k})}$ | $var(\hat{S}(t))$ |
|:--------|:------- |:--------|:--------------------------- |:--------------------------------|:------------------------------------------------------------------- | :------------------------------------ |:----------------- |:---------------- |
| 0       | 21      |0        |0                            |1                                | 1                                                                   | 0                                     |0 | 
| 6       | 21      |3        |$\frac{3}{21}$=```r round(3/21, 3) ```  |$(1-\frac{3}{31})$=```r round(3/21, 3)```  |$(1 - \frac{3}{31})$=```r round(1 - 3/21, 3)```                                | $\frac{3}{21 * 18}$                   | $0.0058$
|$6^*$    | 18      |0        |0                            |1                                |$(1 - \frac{3}{31})$=```r round(1 - 3/21, 3)```                                | $0$                                   | $0.0058$
|7        | 17      |1        |$\frac{1}{17}$ =```r round(1/17, 3) ```|$(1- \frac{1}{17})$=```r round(1/17, 3)``` |$(1 - \frac{3}{31}) \cdot (1 - \frac{1}{17})$=```r round((1 - 3/21)*(1-1/17), 3) ```| $\frac{7}{17(17 - 1)}$                | $0.0078$



## (b)
_In `R`, using the `survfit` function from the package survival, compute and plot the
Kaplan-Meier estimates of S(t) for the two treatment groups._

```{r b_setup, include = FALSE}
library(KMsurv)
library(survival)
data(drug6mp, package = "KMsurv")
Surv(drug6mp$t2, drug6mp$relapse)

treatment     <- survfit(Surv(t2,relapse) ~ 1, 
                        data = drug6mp)
placebo <- survfit(Surv(t1, rep(1, 21)) ~ 1, data = drug6mp)

```

```{r b}

plot(treatment,
     conf.int=FALSE, 
     lty=1, col = "blue")
lines(placebo, lty=2, conf.int=FALSE, col = "red")
legend(20,0.4, bty="n", lty=c(1,2), col = c("blue", "red") , legend=c("Treatment", "Placebo (not the Band!)"))

```


## (c)
_Consider again only the 6-MP patients. Compute (using only your pocket calculator)
the Nelson-Aalen estimate of the cumulative hazard function and its estimated variance._

The Nelson-Aalen Estimator is given as:
\begin{eqnarray*}
\tilde{H}(t)= 
    \begin{cases}  
        0 & \;\mbox{for}\; {t} < t_{1}, \\
        \sum_{t_{k} \leq t} \frac{d_{k}}{n_{k}} & \;\mbox{for}\; {t} \geq t_{1}.
    \end{cases}
\end{eqnarray*} 

The variance of the NA-Estimator is as follows:
$$ \widehat{Var}(\tilde{H}(t)) =  \sum_{t_{k} \leq t} \frac{d_{k}}{n_{k}^2} $$


| $t_{k}$ | $n_{k}$ | $d_{k}$ | $\dfrac{d_{k}}{n_{k}}$                 | $\tilde{H}(t)$                                                 | $\frac{d_{k}}{n_{k}^2}$                      |$\widehat{Var}(\tilde{H}(t)) =  \sum_{t_{k} \leq t} \frac{d_{k}}{n_{k}^2}$   |
|:--------|:------- |:--------|:-------------------------------------- |:---------------------------------------------------------------| :--------------------------------------------|:--------------------------------------------------------------------------- |
| 0       | 21      |0        |0                                       |0                                                               | 0                                            | 0                                                                           | 
| 6       | 21      |3        |$\frac{3}{21}$=```r round(3/21, 3) ```  |$(\frac{3}{21})$=```r round(3/21, 3)```                         | $\frac{3}{21^{2}}$=```r round(3/21^2, 3)```  | $\frac{3}{21^{2}}$=```r round(3/21^2, 3)```                                 |
|$6^*$    | 18      |0        |0                                       |1                                                               | $0$                                          | $\frac{3}{21^{2}}$=```r round(3/21^2, 3)```                                 |
|7        | 17      |1        |$\frac{1}{17}$ =```r round(1/17, 3) ``` |$(\frac{1}{17}) + (\frac{3}{21})$=```r round(1/17 + 3/21, 3)``` | $\frac{1}{17^2}$=```r round(1/17^2, 3)```    |  $\frac{3}{21^{2}} + \frac{1}{17^2}$=```r round(3/21^2 + 1/17^2, 3)```      |



## (d)
_Write a function `nelson.aalen <- function(time,delta)` in R that returns, for a
given set of data, the Nelson-Aalen estimator of the cumulative hazard and its estimated
variance. The argument time corresponds to the event times of the n individuals and
delta is a censoring indicator._

```{r d_setup, include = FALSE}

nelson.aalen <- function(time,delta)
{
  n <- length(time)
  
  # Sort data
  o     <- order(time)
  time  <- time[o]
  delta <- delta[o]
  
  # Frequency table of survival times
  # (censored = 0, event = 1)
  timetable      <- table(time,delta)
  distinct.times <- unique(time)
  
  # Number of events at survival times t_(i)
  d <- timetable[,2]
  
  # Risk set
  timeNoDelta <- apply(timetable, MARGIN=1, sum)
  Y           <- n - c(0,cumsum(timeNoDelta)[-length(timeNoDelta)])
  
  # Nelson-Aalen estimator and standard errors
  na  <- cumsum(d/Y)
  std <- sqrt(cumsum(d/Y^2))
  
  # Return results
  erg <- list(time = distinct.times,
              cumhaz = na,
              std = std)
  return(erg)
}


delta <- drug6mp$relapse
time  <- drug6mp$t2

NAEstimator <- nelson.aalen(drug6mp$t2,delta)


alpha <- 0.05
ci <- cbind(NAEstimator$cumhaz,NAEstimator$cumhaz) + outer(qnorm(1-alpha/2)*NAEstimator$std, c(-1,1))

```

```{r d}
NAEstimator
```


## (e)
_For the 6-MP group, compute and plot in R the Nelson-Aalen estimator along with
95% pointwise asymptotic confidence intervals_

```{r e}
matplot(NAEstimator$time, cbind(NAEstimator$cumhaz,ci),
        xlab="time", ylab="value",
        main="NA-Estimator at alpha =0.05 pointwise CIs", 
        col=c(1,"black","black"), lty=c(1,2,2),lwd=c(3,1,1),type="s")
```


## (f)
<!-- _Graphically compare in R the Nelson-Aalen estimator $H_{NA}(t)$ with the -->
<!-- Breslow estimator $H_B(t) = − ln(\hat S_{KM}(t))$._ -->

```{r f_setup, include=F}
km <- survfit(Surv(t2,relapse)~1, data=drug6mp,type="kaplan-meier",conf.type="none")
breslow <- -log(km$surv)

```


```{r f}
plot(NAEstimator$time,NAEstimator$cumhaz,xlab="t", lty=1, type="s")
lines(km$time,breslow,type="s", lty = 2, col="red")
legend("bottomright",c("Breslow","Nelson-Aalen"),lty=1,col=c("black","red"))

```


# Exercise 2

_Let T be a non-negative continuous random variable denoting the time until the event of
interest. Show that the mean survival time, $\mu = E(T)$, is_ 

$$ \mu = \int_{0}^{\infty} S(t) dt.$$ 



Given that:
$$ \mu = E[T] = \int_{-\infty}^{\infty} t \cdot f_{T}(t) dt = \int_{0}^{\infty} t \cdot f_{T}(t) dt $$
And knowing that time can be denoted as:
$$ t = \int_{0}^{t} dx $$

We now can, for any $y > 0$, see that:
\begin{equation} 
    \begin{aligned} 
      \int_{t = 0}^{t = n} t \cdot f_{T}(t) dt &= \int_{t = 0}^{t = n}  ( \int_{x = 0}^{x = t} ) f_{T}(t) dt \\
                                               &= \int_{t = 0}^{t = n}  \int_{x = 0}^{x = t} f_{T}(t) \cdot dx \cdot dt \\
                                               &= \int_{x = 0}^{x = n}  \int_{t = x}^{t = n}  f_{T}(t) \cdot dt \cdot dx \\
                                               &= \int_{x = 0}^{x = n} ( F_{T}(n) - F_{T}(x)  ) dx 
    \end{aligned}
\end{equation} 

Now assuming that $n \rightarrow \infty$; and furthermore $F_{T}(n) = 1$, we can show that:
$$ E[T] = \int_{0}^{\infty} (1 - F_{T}(x)) dx = \int_{0}^{\infty} f_{T}(x) dx   $$ 
and therefore we can say that:
$$ \mu = \int_{0}^{\infty} S(t) dt$$
is true.


_Given the Kaplan-Meier estimator $\hat{S}_{KM}(t)$ for $S(t)$, use (1) to think of a suitable estimator
$\hat{\mu}$ for ${\mu}$. You may use your proposed estimator to estimate the mean survival time for the
6-MP patients in exercise 1._

\begin{equation} 
    \begin{aligned} 
      \hat{\mu} &= \int_{0}^{\infty} S(t) dt \\
                &= \int_{0}^{\infty} \hat{S}_{KM}(t) dt\\
                &= \sum_{K} \hat{S}_{KM}(t_{k - 1}) \cdot (t_{k} - t_{k - 1}) \\
                &= 1 \cdot 6 + 0.851 \cdot (7 - 6) + 0.807 \cdot (10-7) + 0.753 \cdot (13 - 10) \\
                & + 0.670 \cdot (16 - 13) + 0.627 \cdot (22 - 16) + 0.537 \cdot (23 - 22) + 0.448 \cdot (35 - 23) \\
                &= 23.216
    \end{aligned}
\end{equation} 

23.2 weeks are estimated. Or in full weeks, 24 weeks.

# Exercise 3:
_It follows from the definition of the hazard rate of a continuous random variable $T$ that
$$ P(t ≤ T < t + ∆t | j T ≥ t) = h(t) \Delta t + o(\Delta t), $$
where $∆t$ is a small time interval and $o(∆t)=∆t$ ! 0 as $∆t$ ! 0. Hence, for small $∆t$,
$$ P(t ≤ T < t + ∆t | j T ≥ t) ≈ h(t)∆t =: P_1 :$$
Compare the probability $P(t ≤ T < t + ∆t | j T ≥ t) =: P_2$ with its approximation $P_1$, by
giving an expression for its difference $δ = P_1 − P_2$, in the following two cases:_

## (a)
_For: $T \sim \mathcal{E}(\lambda)$_
We do know that the Exponential distribution is a special case of the Weibull at $\mathcal{WB}(\alpha = 1, \lambda).$
For $P_{1}$ we simply know that:
$$ P_{1} = h(t) \cdot \Delta t = \lambda \cdot \Delta t.$$

For $P_{2}$ we can show that:
\begin{equation} 
    \begin{aligned} 
      P_{2} =& 1 - \dfrac{exp \{ -\lambda (t + \Delta t)\} }{ exp \{ -\lambda t\} } \\
             & 1 - \dfrac{exp \{ -\lambda \cdot t - \lambda \cdot \Delta t)\} }{ exp \{ -\lambda t\} } \\
             & 1 - exp \{ -\lambda \Delta t \} 
    \end{aligned}
\end{equation} 

## (b)
_For $T \sim \mathcal{WB}(\alpha , \lambda)$, with $\alpha = 2$._

We do know that for a $\alpha = 2$ the Weibull distribution takes on the form of an Rayleigh distribution.

For $P_{1} = h(t) \cdot \Delta t = \lambda \cdot \alpha (\lambda \cdot t)^{\alpha - 1} \cdot \Delta t$

\begin{equation} 
    \begin{aligned} 
      P_{1} = h(t) \cdot \Delta t =& \lambda \cdot \alpha (\lambda \cdot t)^{\alpha - 1} \cdot \Delta t \\
             & 2 \cdot \lambda \cdot \lambda \cdot t \cdot \Delta t \\
             & 2 \lambda^{2} \cdot t \cdot \Delta t
    \end{aligned}
\end{equation} 


For $P_{2}$:
\begin{equation} 
    \begin{aligned} 
      P_{2} =& 1 - \dfrac{exp \{ - (\lambda \cdot (t + \Delta t))^{2} \} }{ exp \{ -(\lambda t)^{2}  \} } \\
             & 1 - \dfrac{exp \{ -\lambda^{2}  \cdot (t + \Delta t)^{2}  \} }{ exp \{ -\lambda^{2} \cdot t^{2}  \} } \\
             & 1 - exp \{ -\lambda^{2} ( (t + \Delta t)^{2} - t^{2}) \} 
    \end{aligned}
\end{equation} 

```{r 3_setup, include =F}

delta_t <- seq(0,2,length=1000)       # Grid for delta_t = DELTA t (small time interval)
lambda <- 1

P1 <- lambda*delta_t                  # approximation
P2 <- 1-exp(-lambda*delta_t)          # accurate probability
difference <- P1-P2                  # difference (independent of t)

```

```{r 3}

par(mfrow=c(1,2))
plot(delta_t,P1,type="l",ylab="P( t <= T < t+delta_t | T >= t )",col="blue",main="Exponential distribution")
lines(delta_t,P2,col="magenta")
legend("topleft",legend=c("P_2","P_1 (approximation)"),col=c("magenta","blue"),bg="white",lty=1)
plot(delta_t,abs(difference),type="l",ylab="difference",col="green",main="Difference")
```






