---
site: bookdown::bookdown_site
output:
  pdf_document: default
  bookdown::pdf_book:
    keep_tex: yes
  bookdown::gitbook:
    lib_dir: book_assets
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)

```

# Analysis of Time-to-Event Data: Study Sheet 2

# Submission: René-Marcel Kruse, Doctorial Candidate


# Exercise 1:
_Freireich et al. (1963) report the results of a clinical trial of a drug 6-mercaptopurine (6-
MP) versus a placebo in 42 children with acute leukemia. The trial was conducted at 11
American hospitals. Patients were selected who had a complete or partial remission of their
leukemia induced by treatment with the drug prednisone. (A complete or partial remission
means that either most or all signs of disease had disappeared from the bone marrow.)
The trial was conducted by matching pairs of patients at a given hospital by remission
status (complete or partial) and randomizing within the pair to either a 6-MP or placebo
maintenance therapy. Patients were followed until their leukemia returned (relapse) or until
the end of study. The data are given in the data frame drug6mp in the R package `KMsurv`. Make
use of `help(drug6mp)` to become acquainted with the data set. Note that the remission times
in Freireich et al. (1963) were measured in weeks and not, as indicated in `help(drug6mp)`, in
months!_

## (a)
_Consider only the 6-MP patients. Compute (using only your pocket calculator) the
Kaplan-Meier estimate of the survivor function along with the standard errors._ 

Kaplan-Meier Estimator given as:
$$\hat{S}(t) = \prod_{t_{k} \leq t} (1 - \dfrac{d_{k}}{n_{k}})$$
Variance of KP Estimator is as follows:
$$ var(\hat{S}(t)) = \hat{S}(t)^{2} \cdot \sum_{t_{k} \leq t} \dfrac{d_{k}}{n_{k}(n_{k} - d_{k})})$$

| $t_{k}$ | $n_{k}$ | $d_{k}$ | $\dfrac{d_{k}}{n_{k}}$      | $(1 - \dfrac{d_{k}}{n_{k}})$    | $\hat{S}(t)$                                                        | $\dfrac{d_{k}}{n_{k}(n_{k} - d_{k})}$ | $var(\hat{S}(t))$ |
|:--------|:------- |:--------|:--------------------------- |:--------------------------------|:------------------------------------------------------------------- | :------------------------------------ |:----------------- |:---------------- |
| 0       | 21      |0        |0                            |1                                | 1                                                                   | 0                                     |0 | 
| 6       | 21      |3        |$\frac{3}{21}$=```r round(3/21, 3) ```  |$(1-\frac{3}{31})$=```r round(3/21, 3)```  |$(1 - \frac{3}{31})$=```r round(1 - 3/21, 3)```                                | $\frac{3}{21 * 18}$                   | $0.0058$
|$6^*$    | 18      |0        |0                            |1                                |$(1 - \frac{3}{31})$=```r round(1 - 3/21, 3)```                                | $0$                                   | $0.0058$
|7        | 17      |1        |$\frac{1}{17}$ =```r round(1/17, 3) ```|$(1- \frac{1}{17})$=```r round(1/17, 3)``` |$(1 - \frac{3}{31}) \cdot (1 - \frac{1}{17})$=```r round((1 - 3/21)*(1-1/17), 3) ```| $\frac{7}{17(17 - 1)}$                | $0.0078$



## (b)
_In `R`, using the `survfit` function from the package survival, compute and plot the
Kaplan-Meier estimates of S(t) for the two treatment groups._

```{r b_setup, include = FALSE}
library(KMsurv)
library(survival)
data(drug6mp, package = "KMsurv")
Surv(drug6mp$t2, drug6mp$relapse)

treatment     <- survfit(Surv(t2,relapse) ~ 1, 
                        data = drug6mp)
placebo <- survfit(Surv(t1, rep(1, 21)) ~ 1, data = drug6mp)

```

```{r b}

plot(treatment,
     conf.int=FALSE, 
     lty=1, col = "blue")
lines(placebo, lty=2, conf.int=FALSE, col = "red")
legend(20,0.4, bty="n", lty=c(1,2), col = c("blue", "red") , legend=c("Treatment", "Placebo (not the Band!)"))

```


## (c)
_Consider again only the 6-MP patients. Compute (using only your pocket calculator)
the Nelson-Aalen estimate of the cumulative hazard function and its estimated variance._

The Nelson-Aalen Estimator is given as:
\begin{eqnarray*}
\tilde{H}(t)= 
    \begin{cases}  
        0 & \;\mbox{for}\; {t} < t_{1}, \\
        \sum_{t_{k} \leq t} \frac{d_{k}}{n_{k}} & \;\mbox{for}\; {t} \geq t_{1}.
    \end{cases}
\end{eqnarray*} 

The variance of the NA-Estimator is as follows:
$$ \widehat{Var}(\tilde{H}(t)) =  \sum_{t_{k} \leq t} \frac{d_{k}}{n_{k}^2} $$


| $t_{k}$ | $n_{k}$ | $d_{k}$ | $\dfrac{d_{k}}{n_{k}}$                 | $\tilde{H}(t)$                                                 | $\frac{d_{k}}{n_{k}^2}$                      |$\widehat{Var}(\tilde{H}(t)) =  \sum_{t_{k} \leq t} \frac{d_{k}}{n_{k}^2}$   |
|:--------|:------- |:--------|:-------------------------------------- |:---------------------------------------------------------------| :--------------------------------------------|:--------------------------------------------------------------------------- |
| 0       | 21      |0        |0                                       |0                                                               | 0                                            | 0                                                                           | 
| 6       | 21      |3        |$\frac{3}{21}$=```r round(3/21, 3) ```  |$(\frac{3}{21})$=```r round(3/21, 3)```                         | $\frac{3}{21^{2}}$=```r round(3/21^2, 3)```  | $\frac{3}{21^{2}}$=```r round(3/21^2, 3)```                                 |
|$6^*$    | 18      |0        |0                                       |1                                                               | $0$                                          | $\frac{3}{21^{2}}$=```r round(3/21^2, 3)```                                 |
|7        | 17      |1        |$\frac{1}{17}$ =```r round(1/17, 3) ``` |$(\frac{1}{17}) + (\frac{3}{21})$=```r round(1/17 + 3/21, 3)``` | $\frac{1}{17^2}$=```r round(1/17^2, 3)```    |  $\frac{3}{21^{2}} + \frac{1}{17^2}$=```r round(3/21^2 + 1/17^2, 3)```      |



## (d)
_Write a function `nelson.aalen <- function(time,delta)` in R that returns, for a
given set of data, the Nelson-Aalen estimator of the cumulative hazard and its estimated
variance. The argument time corresponds to the event times of the n individuals and
delta is a censoring indicator._

```{r d_setup, include = FALSE}

nelson.aalen <- function(time,delta)
{
  n <- length(time)
  
  # Sort data
  o     <- order(time)
  time  <- time[o]
  delta <- delta[o]
  
  # Frequency table of survival times
  # (censored = 0, event = 1)
  timetable      <- table(time,delta)
  distinct.times <- unique(time)
  
  # Number of events at survival times t_(i)
  d <- timetable[,2]
  
  # Risk set
  timeNoDelta <- apply(timetable, MARGIN=1, sum)
  Y           <- n - c(0,cumsum(timeNoDelta)[-length(timeNoDelta)])
  
  # Nelson-Aalen estimator and standard errors
  na  <- cumsum(d/Y)
  std <- sqrt(cumsum(d/Y^2))
  
  # Return results
  erg <- list(time = distinct.times,
              cumhaz = na,
              std = std)
  return(erg)
}


delta <- drug6mp$relapse
time  <- drug6mp$t2

NAEstimator <- nelson.aalen(drug6mp$t2,delta)


alpha <- 0.05
ci <- cbind(NAEstimator$cumhaz,NAEstimator$cumhaz) + outer(qnorm(1-alpha/2)*NAEstimator$std, c(-1,1))

```

```{r d}
NAEstimator
```


## (e)
_For the 6-MP group, compute and plot in R the Nelson-Aalen estimator along with
95% pointwise asymptotic confidence intervals_

```{r e}
matplot(NAEstimator$time, cbind(NAEstimator$cumhaz,ci),
        xlab="time", ylab="value",
        main="NA-Estimator at alpha =0.05 pointwise CIs", 
        col=c(1,"black","black"), lty=c(1,2,2),lwd=c(3,1,1),type="s")
```


## (f)
<!-- _Graphically compare in R the Nelson-Aalen estimator $H_{NA}(t)$ with the -->
<!-- Breslow estimator $H_B(t) = − ln(\hat S_{KM}(t))$._ -->

```{r f_setup, include=F}
km <- survfit(Surv(t2,relapse)~1, data=drug6mp,type="kaplan-meier",conf.type="none")
breslow <- -log(km$surv)

```


```{r f}
plot(NAEstimator$time,NAEstimator$cumhaz,xlab="t", lty=1, type="s")
lines(km$time,breslow,type="s", lty = 2, col="red")
legend("bottomright",c("Breslow","Nelson-Aalen"),lty=1,col=c("black","red"))

```





