---
site: bookdown::bookdown_site
output:
  pdf_document: default
  bookdown::pdf_book:
    keep_tex: yes
  bookdown::gitbook:
    lib_dir: book_assets
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)

```

# Analysis of Time-to-Event Data: Study Sheet 4, 
# Submission by: René-Marcel Kruse

# Exercise 1:
_Recall the data file melanoma.dat that was analysed previously. In R, one can use the
survdiff() function from the package survival to test whether male and female patients
have the same survival function._

```{r setup_1, include = FALSE}
library(survival)
melanoma <- read.csv("melanoma.dat", sep="")
delta <- 1*(melanoma$state==1)
```

## (a)
_Create a graph of the Kaplan-Meier estimator stratified by gender along with
95% pointwise confidence bounds using the complementary log-log transformation
(log-log). What do you conclude from this plot regarding the equality of the genderspecific survivor functions?_

## Answer:
```{r ex1_a, echo=FALSE}

melaSurv.maenner <- Surv(melanoma$time[melanoma$sex==1],delta[melanoma$sex==1])
km.maenner <- survfit(melaSurv.maenner~1,type="kaplan-meier",conf.type="log-log")

melaSurv.frauen <- Surv(melanoma$time[melanoma$sex==0],delta[melanoma$sex==0])
km.frauen <- survfit(melaSurv.frauen~1,type="kaplan-meier",conf.type="log-log")

# Plot
#---------------------
par(mfrow=c(1,1))
plot(km.maenner, mark.time=F, col="blue")
par(new=TRUE)
plot(km.frauen, mark.time=F, col="red", axes = FALSE)
legend("bottomright", c("women","men"), col=c("red","blue"),lty=1)
```


## (b)
_Conduct both a log-rank test and a Wilcoxon test to test the null hypothesis that there
is no difference in the survivor functions for men and women. Interpret the obtained
output!_

## Answer:
```{r ex1_b}
survdiff(Surv(melanoma$time,delta) ~ melanoma$sex, rho=0)
survdiff(Surv(melanoma$time,delta)~ melanoma$sex, rho=1)
```


## (c)
_Create a plot of the estimated Kaplan-Meier estimate of the survivor function for
male patients with $95\%$ pointwise confidence bounds using the complementary log-log transformation together with the survivor function of an exponentially distributed
random variable $T \sim \mathcal{E}(0.0002)$. Use the log-rank test to test the null hypothesis that
the distribution of the survival time of men is exponential with $\lambda = 0.0002$. Hint: Make
use of `help(survdiff)` to find out how a one sample test can be performed._

## Answer:
```{r ex1_c, echo=F}

# Plot
#-------
par(mfrow=c(1,1))
plot(km.maenner,mark.time=F,col="blue")
lines(km.maenner$time,exp(-0.0002*km.maenner$time),type="s",col="red")
legend("bottomright", c("random","men"), col=c("red","blue"),lty=1)

# Log-rank test with offset S_0(t) = exp(-lambda_0*time) and lambda_0 = 0.0002
## Offset to fix parameter
#-------------------------------------------------------------------------------
log.rank2 <- survdiff(melaSurv.maenner ~ offset(exp(-0.0002*km.maenner$time)),rho=0)
log.rank2
```


# Exercise 2:
_An approximation to the log-rank test statistic $W_L = U_L^2=V_L$ that was introduced in the
lecture for comparing two survivor curves, and which avoids computing the variance $V_L$, is
as follows:_ 
$$
X^2 = \sum_{k=1}^{2} \dfrac{(O_k - E_k)^2}{E_k} \sim \chi_1^2,
$$
_where $O_k = \sum_{j=1}^{r} d_{kj}$ and $E_k = \sum_{j=1}^{r} e_{kj}$ denotes, for group $k (k = 1, 2)$, the sum of the observed and expected counts over all $r$ distinct failure times across the two groups,
respectively. Using the approximation formula (1), carry out a log-rank test for the breast
cancer data example on slide 2 of the set of slides "Nonparametric methods for comparing
survival distributions". Hint: You may use some of the results given on slide 9. Compare your
results with the ones obtained in the lecture using the test statistic $W_L$._

## Answer:
We do know that the observed are denoted as:
$$ O_k = \sum_{j=1}^{r} d_{kj}, $$
and the expected are denoted as:
$$ E_k = \sum_{j=1}^{r} \dfrac{n_{kj}\cdot d_{j}}{n_j}.$$

The Log-Rank Test results for the underlying data is as follows:
$$ O_1 = \sum_{j=1}^{r} d_{1,j} = d_{1,1} + d_{1,2} + \dots + d_{1,25} = 5$$
$$ O_2 = \sum_{j=1}^{r} d_{2,j} = d_{2,1} + d_{2,2} + \dots + d_{2,25} = 21$$
$$E_1 = \sum_{j=1}^{r} \dfrac{n_{1j}\cdot d_{j}}{n_1} = \dfrac{n_{1,1}\cdot d_{1}}{n_1} + \dfrac{n_{1,2}\cdot d_{2}}{n_2} + \dots + \dfrac{n_{1,25}\cdot d_{25}}{n_{25}} = 9.5652 $$
We can calculate the value of $E_2$ based on the imposed relationship of $O_1 + O_2 = E_1 + E_2$. Following this notation, we can rearrange the relationship as follows:
\begin{align*}
  O_1 + O_2 &= E_1 + E_2 \\
        E_2 &= O_1 + O_2 - E_1 \\
        E_2 &= 5 + 21 - 9.5652 \\
        E_2 &= 16.4348
\end{align*}

Now we only have to insert the calculated values into the original equation for $X^2$.

\begin{align*}
  X^2 &= \sum_{k=1}^{2} \dfrac{(O_k - E_k)^2}{E_k}\\
  X^2 &= \dfrac{(5 - 9.5652)^2}{9.5652} + \dfrac{(21 - 16)^2}{16.4348} \\
  X^2 &= 3.4469
\end{align*}

Resulting in a approximated p-value of: $0.0633$. 

\newpage

# Exercise 3:
_Five hundred and ninety-five persons participate in a case control study of the association of
cholesterol and coronary heart disease (CHD). Among them, 300 persons are known to have
CHD and 295 are free of CHD. To find out if elevated cholesterol is significantly associated
with CHD, the investigator decides to control the effects of smoking. The study subjects are
then divided into two strata: smokers and nonsmokers. The following two tables provide the
data._

\textbf{Smokers} 

|                       | with CHD    |Without CHD  | Total   |
|:----------------------|:-----------:|:-----------:|:-------:|
|Elevated Cholesterol   |             |             |         |
|Yes                    |120          |20           |140      |
|No                     |80           |60           |140      |
|Total                  |200          |80           |280      |

\textbf{Nonsmokers}

|                       | with CHD    |Without CHD  | Total   |
|:----------------------|:-----------:|:-----------:|:-------:|
|Elevated Cholesterol   |             |             |         |
|Yes                    |30           |60           |90       |
|No                     |70           |155          |225      |
|Total                  |100          |215          |315      |


_Conduct an appropriate test to judge whether elevated cholesterol is significantly associated
with CHD after adjusting for the effects of smoking._

## Answer:

We have a two strata problem, therefore, we have to use a Stratified Log-Rank Test.
This test can be formulated as follows:
$$
  W_s = \dfrac{(\sum_{k=1}^{S} U_{L,K})^{2}}{\sum_{k=1}^{S} V_{L,K}} \sim \chi(1)^{2}.
$$
Applied to the underlying question at hand we get:
$$
  W_s = \dfrac{( U_{L,smoker} + U_{L,nonsmoker})^{2}}{ V_{L,smoker} + V_{L,nonsmoker}}.
$$

Where as the variable $U_L$ is defined as:
$$
  U_L = \sum_{j=1}^{r} = (d_{1, j} - e_{1,j}) = d_1 - e_1.
$$
Hence we can assume that:

\begin{align*}
  U_{L,smoker} &= d_{1, smoker} - e_{1, smoker} \\
               &= d_{CHD, smoker} - e_{CHD, smoker} \\
               &= 120 - (\dfrac{200 \cdot 140}{280}) \\
               &= 120 - 100 \\
               &= 20.
\end{align*}

Simillary we can assume that:
$$ 
   U_{L,nonsmoker} = 30 - \dfrac{100 \cdot 90}{315} = 1.429
$$

To calculate $V_{L, smoker}$ we do impose the following equation:

\begin{align*}
  V_{L, smoker} &= \dfrac{n_{CHD, smoker} \cdot n_{No CHD, smoker} \cdot d_{smoker} \cdot (n_{smoker} - d_{smoker})}{n_{smoker}^2 \cdot (n_{smoker} - 1)}  \\
                &= \dfrac{200 \cdot 80 \cdot 140 \cdot (280 - 140)}{280^2 \cdot (280 - 1)} \\
                &= 14.337.
\end{align*}

$V_{L, Non-smoker}$ can be calculated via:

\begin{align*}
  V_{L, non-smoker} &= \dfrac{90 \cdot 225 \cdot 100 (315 - 100)}{315^2 \cdot (315 -1)} \\
                    &= 13.974.
\end{align*}

Now plugging in the calculated values into $W_S$ gives:

$$
  W_S = \dfrac{(20 + 1.429)^2}{14.337 + 13.974} = 16.22 \sim \chi(1)^2
$$

Using this result we arrive at a approximated p-value of $5e^{-}5$


\newpage

# Exercise 4:
_Consider two groups with sizes $n1 = n2 = 100$ from a population of size $n = n1 +n2$. Suppose
that the true survival times of the first group are distributed as $T_1 \sim \mathcal{WB}(3, 0.928)$ and that the second group has true survival times $T2$ with hazard rate $h_2(t) = 3t^2$. For both groups
we assume that censoring times are independent and identically distributed. To begin with,
execute the command `set.seed(1234)`._

## (a)
_Generate right censored survival times for both groups separately. Assume that the
censoring times in both groups are exponentially distributed with parameter $\lambda = 2=3$.
Use the inverse transform sampling method (exercise 5, study sheet 1) to generate true
survival times of group `2`. Combine your data (`time` = observed survival times, `delta`
= censoring indicator, `group` = group membership) into a data frame._

## Answer:

```{r ex4_a, echo=F}
set.seed(1234)
cens <- rexp(n=200,rate=2/3)
time1 <- rweibull(n=100,shape=3,scale=1/0.928)
time2 <- (-log(1-runif(n=100)))^(1/3)
time <- c(time1,time2)
delta <- 1*(time<=cens)
time <- time*(time<=cens)+cens*(time>cens)
group <- c(rep(1,length(time1)),rep(2,length(time2)))
data <- data.frame(cbind(time,delta,group))
head(data)
```

## (b)
_Use a two-sample test of your choice to test whether the survivor functions of both
groups are identical. Give an interpretation of the test result $(\alpha = 0.05)$_

## Answer:
```{r ex4_b, echo=F}
(log.rank <- survdiff(Surv(time,delta) ~ group, data=data, rho=0))
(wilcoxon <- survdiff(Surv(time,delta) ~ group, data=data, rho=1))
sum(delta)
```

## (c)
_Use a one-sample test of your choice to test whether the distribution of the survival times
T of the whole population are distributed as $T \sim \mathcal{WB}(3, 0.928)$. Give an interpretation
of the test result  $(\alpha = 0.05)$._

## Answer:
```{r ex4_c, echo=F}
log.rank2 <- survdiff(Surv(time,delta) ~ offset(exp(-(0.928*time)^3)),data=data, rho=0)
log.rank2
wilcoxon2 <- survdiff(Surv(time,delta) ~ offset(exp(-(0.928*time)^3)),data=data, rho=1)
wilcoxon2
```


## (d)
_Assume that the theoretical event times of the whole population are distributed as
$T \sim \mathcal{WB}(3, 1)$. Compute the Kaplan-Meier estimator, the theoretical survivor function
and the empirical survivor function for the whole population and visualise them in a
single plot. How do the results obtained in (c) change if the empirical survivor function
is used as an offset?_

## Answer:
```{r ex4_d, echo=F}
emp.Survival <- rep(1,nrow(data))
for(i in 1:(nrow(data)-1))
{
  emp.Survival[i+1] <- (nrow(data) - i)/nrow(data)
}
kaplan.meier <- survfit(Surv(time,delta) ~ 1,data=data,conf.type="none")
head(data)
plot(kaplan.meier,mark.time=F,xlab="time",ylab="Survival")
lines(sort(data$time),emp.Survival,type="s",col=2)
lines(sort(data$time),1-pweibull(sort(data$time),shape=3, scale=1),type="l",col=4)
legend("topright",c("Kaplan-Meier","Empirical","Theoretical"),lty=1,
       col=c(1,2,4),bg="white")
log.rank3 <- survdiff(Surv(time,delta) ~ offset(emp.Survival),data=data, rho=0)
log.rank3
log.rank4 <- survdiff(Surv(time,delta) ~ offset(exp(-(1*time)^3)),data=data, rho=0)
log.rank4
```


## (e)
_For group 2, compute the mean and median lifetime and the probability to survive longer
than the mean and median lifetime. Carry out calculations for both the theoretical data
and the censored data._

## Answer:

### Mean:
```{r ex4_e, echo=F}
data2 <- data[which(data$group==2),]
mean.dat <- mean(data2$time)
cat("Empirical:", mean.dat)
mean.th <- gamma(1+1/3)
cat("Theoretical:", mean.th)
mean.km <- 0.9018
cat("Kaplan-Meier:", mean.km)
```

### Median:
```{r ex4_e2, echo=F}
median.dat <- median(data2$time)
cat("Empirical:", median.dat)
median.th <- log(2)^(1/3)         # [log(2)^(1/alpha)] / lambda
cat("Theoretical:", median.th)
median.km <- 0.9390
cat("Kaplan-Meier:", median.km)
```

### Survival times:
```{r ex4_e3, echo=F}
## Theoretical
survival.mean.th   <- 1 - pweibull(q=mean.th,shape=3,scale=1)
survival.median.th <- 1 - pweibull(q=median.th,shape=3,scale=1)
cat("Theoretical survival times")
cat("Mean:", survival.mean.th)
cat("Median:", survival.median.th)

## Empirical estimator
kaplan.meier2        <- survfit(Surv(time,delta)~1,data=data2)
survival.mean.dat   <- kaplan.meier2$surv[max(which(kaplan.meier2$time<=mean.dat))]
survival.median.dat <- kaplan.meier2$surv[max(which(kaplan.meier2$time<=median.dat))]
cat("Empirical estimator")
cat("Mean:", survival.mean.dat)
cat("Median:", survival.median.dat)
## Kaplan-Meier
survival.mean.km   <- kaplan.meier2$surv[max(which(kaplan.meier2$time<=mean.km))]
survival.median.km <- kaplan.meier2$surv[max(which(kaplan.meier2$time<=median.km))]
cat("Kaplan-Meier")
cat("Mean:", survival.mean.km)
cat("Median:", survival.median.km)
```












