---
site: bookdown::bookdown_site
output:
  pdf_document: default
  bookdown::pdf_book:
    keep_tex: yes
  bookdown::gitbook:
    lib_dir: book_assets
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)

```

# Analysis of Time-to-Event Data: Study Sheet 3

# Submission: René-Marcel Kruse, Doctorial Candidate

# Exercise 1:
_In **R** make use of the function `rgamma()` to generate 250 event times from a Gamma
distribution with shape parameter `1.5` and scale parameter `0.75`. Then, generate censored
observations according to the following censoring schemes:_

```{r ex1_setup, include=FALSE}
library(survival)
set.seed(42)
n         <- 250
shape     <- 1.5
scale     <- 0.75
?rgamma
true.t  <- sort(rgamma(n,shape,scale))
(mean.t   <- shape*scale)
(var.t    <- shape*scale^2)
par(mfrow=c(1,1))
hist(true.t)
abline(v=mean.t,col=2)
abline(v=mean.t+sqrt(var.t),col=2,lty=2)
abline(v=mean.t-sqrt(var.t),col=2,lty=2)
survivor <- 1-pgamma(true.t,shape,scale)
cum.hazard <- -log(survivor)
NA_estimator <- function(time,delta)
{
  n <- length(time)
  
  # Sort data
  o     <- order(time)
  time  <- time[o]
  delta <- delta[o]
  
  # Frequency table of survival times
  # (censored = 0, event = 1)
  timetable      <- table(time,delta)
  distinct.times <- unique(time)
  
  # Number of events at survival times t_(i)
  d <- timetable[,2]
  
  # Risk set
  timeNoDelta <- apply(timetable, MARGIN=1, sum)
  Y           <- n - c(0,cumsum(timeNoDelta)[-length(timeNoDelta)])
  
  # Nelson-Aalen estimator and standard errors
  na  <- cumsum(d/Y)
  std <- sqrt(cumsum(d/Y^2))
  
  # Return results
  erg <- list(time = distinct.times,
              cumhaz = na,
              std = std)
  return(erg)
}
```

## (a)
_For every individual $i = 1, . . . , n$ there is a fixed observation period $c_{i}$ (type $I$ censoring).
Assume that the fixed observation period is the same for all individuals and set
$c = 1, 3, 5$._

## Answer:

**Censor times**
```{r ex1a_1}
t1     <- pmin(true.t,1)
t2     <- pmin(true.t,3)
t3     <- pmin(true.t,5)
```

**deltas**
```{r ex1a_2}
d1 <- 1*(true.t<=1)
d2 <- 1*(true.t<=3)
d3 <- 1*(true.t<=5)
```

**Survival**
```{r ex1a_3}
surv1  <- Surv(t1,d1)
surv2  <- Surv(t2,d2)
surv3  <- Surv(t3,d3)
```

**KM-estimator**
```{r ex1a_4}
km1 <- survfit(surv1 ~ 1,type="kaplan-meier")
km2 <- survfit(surv2 ~ 1,type="kaplan-meier")
km3 <- survfit(surv3 ~ 1,type="kaplan-meier")
```

**Nelson-Ahlen Estimator**
```{r ex1a_5}
na1 <- NA_estimator(t1,d1)
na2 <- NA_estimator(t2,d2)
na3 <- NA_estimator(t3,d3)
```

**Sum of total events**
```{r ex1a_6}
sum(d1)
sum(d2)
sum(d3)
```

## Answer:

```{r ex1a_7, echo=FALSE, fig.asp = .5}
par(mfrow=c(1,2),mar=c(5,5,4,2)+0.1)
plot(km3,col=2,main=paste("Kaplan-Meier estimator"),
     xlab="t",xlim=c(0,5.2))
lines(km2,col=3)
lines(km1,col=4)
abline(v=c(1,3,5),col=c(4,3,2))
lines(true.t[true.t<=5],survivor[true.t<=5],col=1)
legend("topright",c("c=1","c=3","c=5","Model"),lty=1,col=c(4,3,2,1),bg="white")

plot(na3$time,na3$cumhaz,col=2,type="s",main=paste("Nelson-Aalen estimator"),
     xlab="t",xlim=c(0,5.2),ylim=c(0,5))
lines(na2$time,na2$cumhaz,type="s",col=3)
lines(na1$time,na1$cumhaz,type="s",col=4)
abline(v=c(1,3,5),col=c(4,3,2))
lines(true.t[true.t<=5],cum.hazard[true.t<=5],col=1)
legend("topright",c("c=1","c=3","c=5","Model"),lty=1,col=c(4,3,2,1),bg="white")
```


## (b)
_The study continues until the failure of the first $k$ individuals, where $k$ is some
predetermined integer $(k < n)$ (type $II$ censoring). Set $k = 100, 150, 200$._

**Censoring Times**
```{r ex1b_1}
t1 <- c(true.t[1:100], rep(true.t[100], n-100))
t2 <- c(true.t[1:150],rep(true.t[150],n-150))
t3 <- c(true.t[1:200],rep(true.t[200],n-200))
```

**Deltas**
```{r ex1b_2}
d1 <- c(rep(1,100), rep(0,n-100))
d2 <- c(rep(1,150),rep(0,n-150))
d3 <- c(rep(1,200),rep(0,n-200))
```

**Survivals**
```{r ex1b_3}
surv1 <- Surv(t1,d1)
surv2 <- Surv(t2,d2)
surv3 <- Surv(t3,d3)
```

**Kaplan-Meier Estimator**
```{r ex1b_4}
km1 <- survfit(surv1~1,type="kaplan-meier")
km2 <- survfit(surv2~1,type="kaplan-meier")
km3 <- survfit(surv3~1,type="kaplan-meier")
```


**Nelson-Aalen Estimator**
```{r ex1b_5}
na1 <- NA_estimator(t1,d1)
na2 <- NA_estimator(t2,d2)
na3 <- NA_estimator(t3,d3)
```

## Answer:

```{r ex1b_7, echo = FALSE, fig.asp = .5}
par(mfrow=c(1,2),mar=c(5,5,4,2)+0.1)
plot(km3,col=2,main=paste("Kaplan-Meier estimator"),
     xlab="t",xlim=c(0,5.2))
lines(km2,col=3)
lines(km1,col=4)
abline(v=c(max(km1$time),max(km2$time),max(km3$time)),col=c(4,3,2))
lines(true.t[true.t<=5],survivor[true.t<=5],col=1)
legend("topright",c("k = 100","k = 150","k = 200","Model"),lty=1,col=c(4,3,2,1),bg="white")

plot(na3$time,na3$cumhaz,col=2,type="s",main=paste("Nelson-Aalen estimator"),
     xlab="t",xlim=c(0,5.2),ylim=c(0,5))
lines(na2$time,na2$cumhaz,type="s",col=3)
lines(na1$time,na1$cumhaz,type="s",col=4)
abline(v=c(max(na1$time),max(na2$time),max(na3$time)),col=c(4,3,2))
lines(true.t[true.t<=5],cum.hazard[true.t<=5],col=1)
legend("topright",c("k = 100","k = 150","k = 200","Model"),lty=1,col=c(4,3,2,1),bg="white")
```

## (c)
_Each unit has associated with it a potential censoring time $C_i$ and a potential lifetime
$T_i$ , which are assumed to the independent random variables (type $III$ censoring or
random censoring). Assume that the censoring times are exponentially distributed with
parameter $\lambda = 0.25, 0.5, 1$._

**Censoring times**

```{r ex1c_1}
c1     <- rexp(n,rate=0.25)
t1     <- pmin(true.t,c1)
c2     <- rexp(n,rate=0.5)
t2     <- pmin(true.t,c1)
c3     <- rexp(n,rate=1)
t3     <- pmin(true.t,c1)
```

**Deltas**

```{r ex1c_2}
d1 <- 1*(true.t<=c1)
d2 <- 1*(true.t<=c2)
d3 <- 1*(true.t<=c3)
```

**Survivals**

```{r ex1c_3}
surv1 <- Surv(t1,d1)
surv2 <- Surv(t2,d2)
surv3 <- Surv(t3,d3)
```

**Sum of total events**

```{r ex1c_4}
sum(d1)
sum(d2)
sum(d3)
```

**KM estimator**

```{r ex1c_5}
km1 <- survfit(surv1~1,type="kaplan-meier")
km2 <- survfit(surv2~1,type="kaplan-meier")
km3 <- survfit(surv3~1,type="kaplan-meier")
```


**NA estimator**

```{r ex1c_6}
na1 <- NA_estimator(t1,d1)
na2 <- NA_estimator(t2,d2)
na3 <- NA_estimator(t3,d3)
```

## Answer:

```{r ex1c_7,echo=FALSE, fig.asp = .5}
par(mfrow=c(1,2),mar=c(5,5,4,2)+0.1)
plot(km3,col=2,main=paste("Kaplan-Meier estimator"),
     xlab="t",xlim=c(0,5.2),mark.time=F)
lines(km2,col=3,mark.time=F)
lines(km1,col=4,mark.time=F)
abline(v=c(max(km1$time),max(km2$time),max(km3$time)),col=c(4,3,2))
lines(true.t,survivor,col=1)
legend("topright",c("lambda = 0.25","lambda = 0.50","lambda = 1.00","Model"),lty=1,col=c(4,3,2,1),bg="white")

plot(na3$time,na3$cumhaz,col=2,type="s",main=paste("Nelson-Aalen estimator"),
     xlab="t",xlim=c(0,5.2),ylim=c(0,5))
lines(na2$time,na2$cumhaz,type="s",col=3)
lines(na1$time,na1$cumhaz,type="s",col=4)
abline(v=c(max(na1$time),max(na2$time),max(na3$time)),col=c(4,3,2))
lines(true.t,cum.hazard,col=1)
legend("topright",c("lambda = 0.25","lambda = 0.50","lambda = 1.00","Model"),lty=1,col=c(4,3,2,1),bg="white")
```


\newpage

# Exercise 2:

_The file `melanoma.dat`, which is stored along with a description in the Stud.IP folder “Data”,
contains data of a clinical study at the Department of Plastic Surgery, University Hospital
of Odense, Denmark. 205 patients were observed during the years 1962 to 1977 after a skin
cancer operation (removal of the tumor and the surrounding skin), either up to the end of
the year 1977 or their early death._

## (a)
_Write a function `lifetable <- function(time,delta,grid){...}` in **R** to imple-
ment the life-table method. The argument time corresponds to the event `times` of the
n individuals, `delta` is a censoring indicator and `grid` contains the points used for
discretization of the time axis. For a given set of data, the function should return the
conditional probability of experiencing the event in each interval, the estimated survival
function at the start of each interval, and the estimated density and hazard rate at the
midpoint of each interval._

```{r ex2a_setup, include=FALSE}
melanoma <- read.csv("melanoma.dat", sep="")
time  = melanoma$time
delta = 1*(melanoma$state==1)
grid  =  seq(0,6000,100)
```

## Answer: ... Fortunately still had the solutions from the year before last
```{r ex2a_1}
lifetable <- function(time,delta,grid)
{
  q <- length(grid)-1 
  n <- length(time)   
  d <- numeric(q)     
  Y <- rep(n,q+1)     
  w <- numeric(q)     
  h <- numeric(q)     
  
  for(k in 1:q)
  {
    d[k] <- sum(delta[time>=grid[k] & time<grid[k+1]])
    w[k] <- length(delta[time>=grid[k] & time<grid[k+1]])-d[k]
    Y[k+1] <- Y[k] - d[k] - w[k]
    h[k] <- grid[k+1] - grid[k]
  }
  cp.event <- survivor <- density <- hazard <- numeric(q)
  cp.event <- d / (Y[-(q+1)] - w/2)
  survivor <- cumprod(1-cp.event)
  density <- ( c(1,survivor[-q]) - survivor ) / h
  hazard <- 2*cp.event / (h * (2-cp.event))
  cp.event <- c(cp.event,cp.event[q])
  survivor <- c(1,survivor)
  density <- c(density,density[q])
  hazard <- c(hazard,hazard[q])
  erg <- list(cp.event=cp.event,
              survivor=survivor,
              density=density,
              hazard=hazard)
  return(erg)
}
```



## (b)
_Apply your function written in (a) to the skin cancer data. Define an appropriate
censoring indicator `delta` and use equidistant time intervals `(grid)` between 0 and
6000 with interval lengths 100, 500, 1000 and 2000. Display each quantity that is asked
for in (a) graphically and interpret the results._

## Answer:

```{r ex2b_1, echo=FALSE}
t <- list(seq(0, 6000, 100), seq(0, 6000, 500),seq(0, 6000, 1000),seq(0, 6000, 2000))
lt <- list(0, 0,0,0)
for (i in 1:4){ lt[[i]] <- lifetable(time, delta, t[[i]])}

par(mfrow=c(2, 2))
plot(t[[1]], lt[[1]]$density, type="s", ylim=c(0, 0.0003))
for (i in 2:4) lines(t[[i]], lt[[i]]$density, type="s", col=i)
title("Density")
legend("topright", c("h=100", "h=500", "h=1000", "h=2000"), col=1:4, lty=1)
plot(t[[1]], lt[[1]]$survivor, type="s", ylim=c(0.6, 1.0))
for (i in 2:4) lines(t[[i]], lt[[i]]$survivor, type="s", col=i)
title("Survivor function")
legend("topright", c("h=100", "h=500", "h=1000", "h=2000"), col=1:4, lty=1)
plot(t[[1]], lt[[1]]$cp.event, type="s", ylim=c(0, 0.3))
for (i in 2:4) lines(t[[i]], lt[[i]]$cp.event, type="s", col=i)
title("Conditional probability of experiencing the event")
legend("topright",  c("h=100", "h=500", "h=1000", "h=2000"), col=1:4, lty=1)
plot(t[[1]], lt[[1]]$hazard, type="s", ylim=c(0, 0.0005))
for (i in 2:4) lines(t[[i]], lt[[i]]$hazard, type="s", col=i)
title("Hazard rate")
legend("topright",  c("h=100", "h=500", "h=1000", "h=2000"), col=1:4, lty=1)
```


## (c)
_In **R**, compute and plot the Kaplan-Meier estimate of the survivor function for the data
of survival after malignant melanoma._

## Answer:

```{r ex2c_1, echo=FALSE, fig.asp = .5}
library(survival)
delta <- 1*(melanoma$state==1)
melaSurv <- Surv(time,delta)
km <- survfit(melaSurv~1,type="kaplan-meier",conf.type="none")
par(mar=c(5, 5, 4, 2) + 0.1)
plot(km, main="KM estimator",xlab="t",)
```


## (d)
_Graphically compare the Kaplan-Meier estimate obtained in (c) with the estimate of
the survivor function obtained by the life-table method._

## Answer:
```{r ex2d_1, echo=FALSE, fig.asp = .5}
par(mfrow=c(1,2),mar=c(5, 5, 4, 2) + 0.1)
plot(km,main="Survival",xlab="t")
a <- seq(0,6000,100)
lt <- lifetable(time,delta,a)
lines(a,lt$survivor,type="s",col="blue")
legend("bottomleft",legend=c("Kaplan-Meier","grid = 100"),col=c(1,"blue"),lty=1)

plot(km,main="Survival",xlab="t")
a <- seq(0,6000,500)
lt <- lifetable(time, delta, a)
lines(a, lt$survivor, type="s",col="blue")
legend("bottomleft",legend=c("Kaplan-Meier","grid = 500"),col=c(1,"blue"),lty=1)
```


```{r ex2d_2, echo=FALSE, fig.asp = .5}
par(mfrow=c(1,2),mar=c(5, 5, 4, 2) + 0.1)
plot(km,main="Survival",xlab="t")
a <- seq(0,6000,1000)
lt <- lifetable(time, delta, a)
lines(a, lt$survivor, type="s",col="blue")
legend("bottomleft",legend=c("Kaplan-Meier","grid = 1000"),col=c(1,"blue"),lty=1)

plot(km,main="Survival",xlab="t")
a <- seq(0,6000,2000)
lt <- lifetable(time, delta, a)
lines(a, lt$survivor, type="s",col="blue")
legend("bottomleft",legend=c("Kaplan-Meier","grid = 2000"),col=c(1,"blue"),lty=1)
```


\newpage

# Exercise 3:
_Several proposals for computing pointwise confidence intervals for values of the survival
function do exist._

## (a)
_Recall the data file `melanoma.dat` that was analysed in exercise 2. In **R**, use the
`survfit()` function from the package survival to compute the Kaplan-Meier estimate
of the survivor function and compare various pointwise $95\%$ confidence intervals that
are available in the option `conf.type`._
```{r ex3a_setup, include=FALSE}
melanoma <- read.csv("melanoma.dat", sep="")
library(survival)
data(melanoma)
delta <- 1*(melanoma$state==1)
time  <- melanoma$time
melaSurv <- Surv(time,delta)
```

## Answer:

```{r ex3a_1, echo=FALSE, fig.asp = .5}
par(mfrow=c(1,1),mar=c(5,5,4,2)+0.1)
plot(km2,mark.time=F,col="blue",lwd=1,xlab="t",ylim=c(0.5,1))
par(new=TRUE)
plot(km3,mark.time=F,col="red",lwd=1,ylim=c(0.5,1))
par(new=TRUE)
plot(km1,mark.time=F,lwd=1,main="Different CIs",ylim=c(0.5,1))
legend("bottomleft",c("standard","log(S(t))","log(-log(S(t)))"),
       col=c("black","blue","red"), lty=1, lwd=2)
```


## (b)
_Use the delta method that was introduced in the lectures to derive the asymptotic
variance of $\mathrm{ln} ( -\mathrm{ln} ( \hat{S}(t)  )$._

## Answer:

The delta method is given as:
$$ Var(f(x)) = Var(x) \cdot \Bigg( \dfrac{\delta(f(x))}{\delta x}  \Bigg)^{2} $$

Applied to our problem we get:

\begin{align}
    \mathrm{Var}\Big( \mathrm{ln} ( -\mathrm{ln} ( \hat{S}(t)  )  \Big) &= \mathrm{Var} \Big(  \mathrm{ln}( \hat{S}(t)  ) \Big) \cdot \Bigg( \dfrac{\delta \big( \mathrm{ln}( -\mathrm{ln}( \hat{S}(t)  )  \big) }{\delta \Big(  \mathrm{ln}( \hat{S}(t)  ) \Big)}  \Bigg)^{2}  \\
    &=  \mathrm{Var} \Big(  \mathrm{ln}( \hat{S}(t)  ) \Big) \cdot \dfrac{1}{\mathrm{ln}( \hat{S}(t) )^{2} } \\
    &= \sum_{t_{i} \leq t} \dfrac{d_{k}}{ n_{k} (n_{k} - d_{k})} \cdot \dfrac{1}{\mathrm{ln}( \hat{S}(t) )^{2} }
\end{align}



## (c)
_Based on the result obtained in (b) give a corresponding asymptotic $100 ( 1 - \alpha ) \%$ 
confidence interval for $S(t)$._

## Answer:

CI for $\hat{S}(t)$ given by:

$$ CI_{\pm} = \mathrm{ln} \big( \mathrm{-ln} (\hat{S}(t) \big) \pm z_{\frac{\alpha}{2}} \cdot \sqrt{  \mathrm{Var}\Big( \mathrm{ln} ( -\mathrm{ln} ( \hat{S}(t)  )  \Big) }$$


\newpage

# Exercise 4:
_In this exercise we consider data on the efficiency of bone marrow transplants for patients
with acute myelotic leukemia (AML) and acute lymphoblastic leukemia (ALL). The data are
reported in section 1.3 and appendix D of Klein and Moeschberger (2003) and details of this
study are given in Copelan et al. (1991). The data are contained in the data frame `bmt` in
the **R** package `KMsurv`. Make use of  `help(bmt)` to become acquainted with the data set._

```{r ex4_setup, include = FALSE}
library(KMsurv)
data(bmt)
library(survival)
library(km.ci)
```

## (a)
_We shall focus on the disease-free survival probabilities for the 38 patients with ALL. An
individual is said to be disease free at a given time after transplant if that individual is
alive without the recurrence of leukemia. In **R**, construct and graphically compare $95\%$
pointwise confidence intervals, Equal-Precision (EP) confidence bands and Hall-Wellner
confidence bands for the disease-free survival function in the range $100 \leq t \leq 600$.
Hint: In the lecture, we used the function `confBands()` from the R package `OIsurv` to
construct confidence bands. Confidence bands can also be computed using the `km.ci()`
function in the **R** package `km.ci`._

```{r ex4a_1, echo=FALSE, fig.asp = .5}
fit <- survfit(Surv(bmt$t2[bmt$group==1],bmt$d3[bmt$group==1]) ~ 1,
               type="kaplan-meier",conf.int=FALSE)
pw <- km.ci(fit, conf.level=0.95, tl=NA, tu=NA, method="log")
epbands <- km.ci(fit, conf.level=0.95, tl=NA, tu=NA, method="logep")
hwbands <- km.ci(fit, conf.level=0.95, tl=NA, tu=NA, method="loghall")
par(mfrow=c(1,1),mar=c(5,5,4,2)+0.1)
plot(fit,mark.time=F,col="black",xlab="Days",ylab="Estimated survival function", lty=1, main="95%-CI ", xlim=c(100,600))
lines(pw, mark.time=F, col="blue", lty=1)
lines(epbands, mark.time=F, col="red", lty=1)
lines(hwbands, mark.time=F, col="black", lty=1)

legend("bottomleft",c("Pointwise CI","EP CB","Hall-Wellner CB","KM estimate"),
       col=c("blue","red","black","black"), lty=c(2,2,2,1), lwd=1)
```

## (b)
_Estimate the mean disease-free survival time for the three disease groups._

```{r ex4b_1, echo=FALSE, fig.asp = .5}
bmtSurv  <- Surv(bmt$t2[bmt$group==1],bmt$d3[bmt$group==1])
bmtsurv2 <- Surv(bmt$t2[bmt$group==2],bmt$d3[bmt$group==2])
bmtsurv3 <- Surv(bmt$t2[bmt$group==3],bmt$d3[bmt$group==3])
print(survfit(bmtSurv~1), print.rmean=TRUE)
print(survfit(bmtsurv2~1), print.rmean=TRUE)
print(survfit(bmtsurv3~1), print.rmean=TRUE)

### Median survival time and CI
plot(survfit(bmtSurv~1))
abline(h = 0.5)
abline(v = quantile(survfit(bmtSurv~1), probs = 0.5)$quantile)
abline(v = quantile(survfit(bmtSurv~1), probs = 0.5)$lower)
```

## (c)
_Estimate the median disease-free survival time for the three disease groups and
construct a $95\%$ confidence interval for the median._

```{r ex4c_1, echo=FALSE, fig.asp = .5}
quantile(survfit(bmtSurv~1), probs = 0.5)$quantile
quantile(survfit(bmtSurv~1), probs = c(0.025, 0.5, 0.975))$quantile
quantile(survfit(bmtsurv2~1), probs = c(0.025, 0.5, 0.975))$quantile
quantile(survfit(bmtsurv3~1), probs = c(0.025, 0.5, 0.975))$quantile
```