---
site: bookdown::bookdown_site
output:
  pdf_document: default
  bookdown::pdf_book:
    keep_tex: yes
  bookdown::gitbook:
    lib_dir: book_assets
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
library(survival)
```

# Analysis of Time-to-Event Data: Study Sheet 8, 
# Submission by: René-Marcel Kruse 

# Exercise 1:
_Let T denote survival time with survival function $S_T(t)$. Simulate a sample of size $n = 1500$ from a Cox model with hazard rate._
$$
  h(t; x) = t \cdot exp \{ 0.5 x \}. 
$$
_Use the inverse transform sampling method developed in exercise 5, study sheet 7. Simulate the covariate $x_1$ from a uniform distribution on the interval `[-3; 3]` and the censoring times from a uniform distribution on the interval `[0; 6]`. Plot the Cox-Snell residuals against the cumulative hazard rate to check the overall goodness-of-fit of the fitted model. For the derivation of the distribution of the Cox-Snell residuals use the distribution of $Y = - ln(S_T (T)) \sim E(\lambda = 1)$_

## Answer:

### Draw samples:

```{r ex1_paras}
x <- runif(1500, -3, 3)
censortimes <- runif(1500, 0, 6)
u <- runif(1500)
``` 

### Insverse sampling method:

```{r ex1_inv}
samples <- sqrt(-2 * log(1 - u) * exp(-0.5 * x))
``` 

### Event times and censoring:

```{r ex1_event}
time <- pmin(samples, censortimes)
delta <- (samples <= censortimes) * 1
``` 

### Cox Model:
```{r ex1_cox}
coxmodel <- coxph(Surv(time, delta) ~ x)
summary(coxmodel)
``` 

### Cox-Snell residuals:

```{r ex1_resids, echo=FALSE}
res.cs <- delta - residuals(coxmodel, type="martingale")
res.surv <- survfit(Surv(res.cs,delta) ~ 1)
res.cumhaz <- -log(res.surv$surv)

plot(res.surv$time,res.cumhaz,xlab="Residuals",ylab="Cum. Hazard",
     type="s", col=2, main="Cox-Snell Residuals Model fit")
lines(res.surv$time, res.surv$time, lty=4)
``` 

# Exercise 2:
_Simulate a sample of size $n = 1500$ from a Cox model with hazard rate_

  $$
  \text{(a)} \quad  h(t; x) = t \cdot exp \{ sin(x_1) + 0.5x_2 \}
  $$

  $$
  \text{(b)} \quad  h(t; x) = t \cdot exp \{ x_1^2 + 0.5x_2 \}
  $$

_Use the inverse transform sampling method developed in exercise 5, study sheet 7. Simulate the covariate $x_1$ and $x_2$ from a uniform distribution on the interval `[-3; 3]` and the censoring times from a uniform distribution on the interval `[0; 6]`. Obtain the martingale residuals and deviance residuals and check whether one can use them to make conclusions about the functional form of the covariate $x_1$. The `loess()` function can be used to smooth the residuals._

## Answer:

```{r ex21_paras}
x1 <- runif(1500, -3, 3)
x2 <- runif(1500, -3, 3)
censortimes <- runif(1500, 0, 6)
u <- runif(1500)
``` 

### (a):

### Insverse sampling method:

```{r ex21_inv}
samples <- sqrt(-2 * log(u) * exp(-sin(x1)-0.5*x2))
``` 

### Event times and censoring:

```{r ex21_event}
time <- pmin(samples, censortimes)
delta <- (samples <= censortimes) * 1
``` 

### Cox Model:
```{r ex21_cox}
coxmodel <- coxph(Surv(time, delta) ~ x2)
summary(coxmodel)
``` 

### Martingale Residuals:

```{r ex21_mart}
m.residuals <- residuals(coxmodel, type="martingale")
m.loess1 <- loess(m.residuals ~ x1, degree=1)
m.loess2 <- loess(m.residuals ~ x1, degree=2)
``` 

```{r ex211_mart_plot, echo=FALSE}
plot(x1, m.residuals, pch=".", col=1, main="Martingale residuals", ylim = c(-6, 6))
```

```{r ex212_mart_plot, echo=FALSE}
par(mfrow=c(1,2))
plot(x1, predict(m.loess1), ylim= c(-6, 6), 
     ylab="m.residuals", main="Fit with loess degree=1")
points(x1, m.residuals, pch=".", col=3)
plot(x1, predict(m.loess2), ylim= c(-6, 6), 
     ylab="m.residuals", main="Fit with loess degree=2")
points(x1, m.residuals, pch=".", col=2)
``` 

### Deviance Residuals:

```{r ex21_devi}
d.residuals <- residuals(coxmodel, type="deviance")
d.loess <- loess(d.residuals ~ x1, degree=2)
``` 

```{r ex21_devi_plot, echo=FALSE}
par(mfrow=c(1,2))
plot(x1, d.residuals, pch=".", col=2, main="Deviance residuals", ylim= c(-6, 6))
plot(x1, predict(d.loess), ylim= c(-6, 6),
     ylab="d.residuals", main="Fit with loess")
points(x1, d.residuals, pch=".", col=2)
``` 


### (b):

### Insverse sampling method:

```{r ex22_inv}
samples <- sqrt(-2 * log(u) * exp(- x1^2 - 0.5 * x2))
``` 

### Event times and censoring:

```{r ex22_event}
time <- pmin(samples, censortimes)
delta <- (samples <= censortimes) * 1
``` 

### Cox Model:
```{r ex22_cox}
coxmodel <- coxph(Surv(time, delta) ~ x2)
summary(coxmodel)
``` 

### Martingale Residuals:

```{r ex22_mart}
m.residuals <- residuals(coxmodel, type="martingale")
m.loess1 <- loess(m.residuals ~ x1, degree=1)
m.loess2 <- loess(m.residuals ~ x1, degree=2)
``` 

```{r ex221_mart_plot, echo=FALSE}
plot(x1, m.residuals, pch=".", col=1, main="Martingale residuals", ylim = c(-6, 6))
```

```{r ex222_mart_plot, echo=FALSE}
par(mfrow=c(1,2))
plot(x1, predict(m.loess1), ylim= c(-6, 6), 
     ylab="m.residuals", main="Fit with loess degree=1")
points(x1, m.residuals, pch=".", col=3)
plot(x1, predict(m.loess2), ylim= c(-6, 6), 
     ylab="m.residuals", main="Fit with loess degree=2")
points(x1, m.residuals, pch=".", col=2)
``` 

### Deviance Residuals:

```{r ex22_devi}
d.residuals <- residuals(coxmodel, type="deviance")
d.loess <- loess(d.residuals ~ x1, degree=2)
``` 

```{r ex22_devi_plot, echo=FALSE}
par(mfrow=c(1,2))
plot(x1, d.residuals, pch=".", col=2, main="Deviance residuals", ylim= c(-6, 6))
plot(x1, predict(d.loess), ylim= c(-6, 6),
     ylab="d.residuals", main="Fit with loess")
points(x1, d.residuals, pch=".", col=2)
``` 

# Exercise 3:

_In the lectures, the martingale property has been stated as follows (see slide 7 of the set of slides "Refinements of the semiparametric proportional hazards model"):_
$$
  E[dM(t)|F_{t-}] = 0 \quad \text{for all }t.
$$
_Show that equation (1) is equivalent to_
$$
  E[M(t)|F_{s}] = M(s) \quad \text{for all }s.
$$

## Answer:

We do know that $M(t)$ is defined as follows
$$
  M(t) = N(t) - \Delta(t)
$$
where as $N(t)$ represents a counting process and $\Delta(t)$ represents the cumulative intensity of the process itself.

We can now employing the law of total expectations reformulate as follows

\begin{align*}
  E[M(t)|F_{s}] - M(s) &= E[M(t) -  M(s)|F_{s}]  \\
                       &= E[\int_{s}^{t} dM(u) | F_s] \\
                       &= \int_{s}^{t} E[d m(u) | F_s] \\
                       &= \int_{s}^{t} E[ E[d M(u) | F_s, F_{u-}] | F_s ] \\
                       &= \int_{s}^{t} E[ E[d M(u) | F_{u-}] | F_s ] \\
                       &= 0
\end{align*}

# Exercise 4:

_The file `resmelanoma.prn` that is available in the Stud.IP folder "Data" contains survival times from 30 resected melanoma patients (for a description of the data, see the file `resmelanomahelp.txt`).1 Let $ageg$ denote the age group with $ageg = 1$ if age $<$ 45 and $ageg = 2$ otherwise. Fit the survival times with an ageg-stratified Cox proportional hazards model with the covariates sex and treatment received._

## Answer:
```{r ex4, echo=FALSE}
library(survival)
data <- read.table("resmelanoma.prn",sep="")

colnames(data) <- c("ID", "AGE", "SEX", "INI2",
                    "INI3A", "INI3B", "INI4A", "TRT", "RTIME", "CRT", "STIME", "CST")
head(data)

data$AGEG <- 2*(data$AGE == 45 | data$AGE > 45)
data$AGEG[data$AGEG == "0"] <- "1"
head(data)


CoxModel.Strat <- coxph(Surv(STIME,CST) ~ SEX + TRT + strata(AGEG), 
                        data=data)
summary(CoxModel.Strat)
``` 


# Exercise 5:
_The file `prison.txt`, which is available in the Stud.IP folder "Data", contains data from an experimental study of recidivism of 432 male prisoners, who were observed for a year after being released from prison. Half of the prisoners were randomly given financial aid when they were released. The following table gives a description of the observed variables:_

| Variable    |   Description                                                                                                        |
|:------------|:---------------------------------------------------------------------------------------------------------------------|
|week         |week of first arrest after release, or censoring time                                                                 |
|arrest       |the event indicator, 1 = arrested , 0 = not                                                                           |
|fin          |1 = received financial aid, 0 = not                                                                                   |
|age          |in years at the time of release                                                                                       |
|race         |1 = black, 0 = others                                                                                                 |
|wexp         |1 = had full-time work experience, 0 = not                                                                            |
|mar          |1 = married, 0 = not                                                                                                  |
|paro         |1 = released on parole, 0 = not                                                                                       |
|prio         |number of prior convictions                                                                                           |
|educ         |codes 2 (grade 6 or less), 3 (grades 6 through 9), 4 (grades 10 and 11), 5 (grade 12), or 6 (some post-secondary)     |
|emp1 - emp52 | 1 = employed in the corresponding week, 0 = not                                                                      |


### (a):

_Fit a Cox model to these data. Use backward selection, which is implemented in the function `stepAIC()` function from the R package MASS, to find the best model according to the Akaike Information Criterion (AIC)_

```{r ex5a, echo=FALSE}
prison <- read.table("prison.txt", header=TRUE)
ex5a <- coxph(Surv(week, arrest) ~ fin + age + race + wexp + mar + paro + prio + as.factor(educ),
            method="efron",data=prison)
library(MASS)
step.AIC<-stepAIC(ex5a,direction="both")
summary(step.AIC)
``` 


### (b):

_In the file prisonlong.txt each row corresponds to one observation per person per week. Fit a Cox model with the time-dependent variable employed to these data._

```{r ex5b, echo=FALSE}
prison.long <- read.table("prisonlong.txt", header=TRUE)

(m2<-coxph(Surv(start, stop, arrest.time) ~ fin + age + prio + mar, data=prison.long))

(m3<-coxph(Surv(start, stop, arrest.time) ~ fin + age + prio + mar + employed, data=prison.long))
``` 


### (c):

_Create a variable employed.lag1 which should contain information whether the person was employed in the previous week. Again, fit a Cox model using the variable employed.lag1 instead to employed._

```{r ex5c, echo=FALSE}
lag1 <- function(x) c(0,head(x,n=-1))
prison.long$employed.lag1 <- unlist(with(prison.long, tapply( employed, subject, lag1)))

(m4<-coxph(Surv(start, stop, arrest.time) ~ fin + age + prio + mar + employed.lag1,data=prison.long))
``` 

### (d):

_How could you check the assumption of proportional hazards for all the variables of the best model found in (a) using interaction terms with time of observation?_

```{r ex5d, echo = FALSE}

(m.fin<- coxph(Surv(start, stop, arrest.time) ~ fin + fin:I(log(stop)) + age + prio + mar + employed, data=prison.long))

(m.age<- coxph(Surv(start, stop, arrest.time) ~ fin + age + age:I(log(stop)) + prio + mar + employed, data=prison.long))

(m.prio<- coxph(Surv(start, stop, arrest.time) ~ fin + age + prio + prio:I(log(stop))+ mar + employed, data=prison.long))

(m.mar<- coxph(Surv(start, stop, arrest.time) ~ fin + age + prio + mar + mar:I(log(stop))  + employed, data=prison.long))

(m.employed<- coxph(Surv(start, stop, arrest.time) ~ fin + age + prio + mar + employed + employed:I(log(stop)), data=prison.long))



```
